---
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->


```{r setup, include=FALSE}
library(emoji)

yaml_metadata <- rmarkdown::yaml_front_matter(here::here(
  "manuscript",
  "manuscript.qmd"
))

title <- yaml_metadata$title
# abstract <- yaml_metadata$abstract

emoji_out <- emoji_glue(":pray::bee:")
```

## Goodness of fit diagnostics for discrete time Markov models of cognitive transitions

::: {.content-visible when-format="markdown"}

- [Cormac Monaghan](https://c-monaghan.github.io/) • Maynooth University
- [Idemauro Antonio Rodrigues de Lara](https://scholar.google.com/citations?user=-f0hcJYAAAAJ&hl=pt-BR) • University of São Paulo
- [Rafael de Andrade Moral](https://rafamoral.github.io/) • Maynooth University
- [Joanna McHugh Power](https://www.maynoothuniversity.ie/people/joanna-mchugh-power) • Maynooth University

:::


---

> Monaghan, C., de Lara, IAR, de Andrade Moral, R, & McHugh Power, J. `r title`.

---

## Abstract

## `r emoji_out`: Note on "hopeful hornet" project name

Because project titles change all the time we used [{codename}](http://svmiller.com/codename/) to generate an generate an [Ubuntu-style](https://wiki.ubuntu.com/DevelopmentCodeNames)
internal-to-us project name that won’t change.


```{r codename, eval=FALSE}
library(codename)

codename_message()
#> code name generated by {codename} v.0.5.0. R version 4.4.2 (2024-10-31 ucrt).

codename(seed = "residual diagnostics", type = "ubuntu")
#> [1] "hopeful hornet"
```

## Licenses

**Text and figures:** All prose and images are licensed under Creative Commons ([CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)).

**Code:** All code is licensed under the [MIT License](LICENSE.md).
